Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
prefinal
minor

Our community builds, improves, and reasons about programming languages. To make
design decisions that benefit users, we need to understand our target language
as well as the real-world needs it answers. Often, we can appeal to our
intuition, as many languages are intended for general purpose programming tasks.
Unfortunately, intuition may fail when looking at domain-specific languages as
these languages were designed for a specific group of users to solve very
specific needs. This is the case of the data science language R.

@@

Dynamic programming languages such as Racket, JavaScript and PHP have been
extended post-facto with various static type systems.  In each case, the
type system was carefully engineered to match the salient characteristics of
the host languages and to foster a particular programming style. For
example, Typed Racket emphasizes functional programming and support the
migration from untyped to fully typed code~\cite{tf-popl08},
Hack~\cite{hack13} and TypeScript~\cite{BAT14} focus on typing
object-oriented features of PHP and JavaScript, respectively. They allow to
intersperse typed and untyped code in a fine-grained manner.

@@

One way to validate the inferred type annotations is to check that different
programs using one of the functions for which a type signature has been inferred
respect the signature. For this purpose, we developed \contractr, an R package
that allows one to decorate function with assertions. We use it to insert type
checks of the arguments and return values. As usual with R, this is not
straightforward.

While \contractr's primary logic has been implemented in C++ to reduce its
runtime overhead, we have not observed a single segfault during our use of
the package.  It has been tested with GNU R-3.5.0 and hardened with a
battery of 400 test cases. It works by modifying function definition to
insert a call to a type-checking function.

>>>>

Our community builds, improves, and reasons about programming languages. To make
design decisions that benefit users, we need to understand the target language
as well as the real-world needs it answers. Often, we can appeal to our
intuition, as many languages are intended for general purpose programming tasks.
Unfortunately, intuition may fail when looking at domain-specific languages as
these languages were designed for a specific group of users to solve very
specific needs. This is the case of the data science language R.

@@

Dynamic programming languages such as Racket, JavaScript and PHP have been
extended post-facto with various static type systems.  In each case, the
type system was carefully engineered to match the salient characteristics of
the host languages and to foster a particular programming style. For
example, Typed Racket emphasizes functional programming and support the
migration from untyped to fully typed code~\cite{tf-popl08},
Hack~\cite{hack13} and TypeScript~\cite{BAT14} focus on typing
object-oriented features of PHP and JavaScript, respectively. They allow to
intersperse typed and untyped code in a fine-grained manner.
Other work looks at retrofitting types onto Lua~\cite{typedlua2014}.

@@

One way to validate the inferred type annotations is to check that different
programs using one of the functions for which a type signature has been inferred
respect the signature. For this purpose, we developed \contractr, an R package
that allows one to decorate function with assertions. We use it to insert type
checks of the arguments and return values.

\contractr's primary logic has been implemented in C++ to reduce its
runtime overhead, and we have not observed a single segfault during our use of
the package.  It has been tested with GNU R-3.5.0 and hardened with a
battery of 400 test cases. It works by modifying function definition to
insert a call to a type-checking function.
